\documentclass{llncs}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb}
\usepackage{hyperref}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{booktabs}
\usepackage{microtype}
% Comment macro — remove the \renewcommand line before submission
\newcommand{\comment}[1]{}

% Lean code styling
\lstdefinelanguage{Lean}{
  keywords={def, theorem, lemma, example, import, open, do, let, if, then, else, for, in, return,
  by, TacticM, Unit, Prop, Type, where, match, with, fun, partial, namespace, end, inductive,
  syntax, initialize, pure, true, false, none, some},
  keywordstyle=\color{blue}\bfseries,
  ndkeywords={getMainGoal, getType, instantiateMVars, evalTactic, throwTacticEx,
    isAppOfArity, getAppArgs, isBVar, isAppOf, registerTraceClass},
  ndkeywordstyle=\color{purple},
  identifierstyle=\color{black},
  sensitive=true,
  comment=[l]{--},
  morecomment=[s]{/-}{-/},
  commentstyle=\color{gray}\itshape,
  stringstyle=\color{orange},
  morestring=[b]",
}

\lstset{
  language=Lean,
  basicstyle=\ttfamily\footnotesize,
  breaklines=true,
  frame=single,
  numbers=left,
  numberstyle=\tiny\color{gray},
  tabsize=2,
  xleftmargin=2em,
}

\title{Generating Lean~4 Tactics from User Specifications:\\
       An Empirical Study of LLM-Assisted Metaprogramming Strategies}

\author{Silvère Gangloff\inst{1} \and Jan Hula\inst{1} \and Miroslav Olšák\inst{2}}

\institute{
  National Supercomputing Centre IT4Innovations, IRAFM,\\
  University of Ostrava, 30.~dubna~22, 70103 Ostrava, Czech Republic\\
  \email{\{silvere.gangloff, jan.hula\}@osu.cz}
  \and
  Czech Institute of Informatics, Robotics and Cybernetics,\\
  Czech Technical University in Prague\\
  \email{miroslav.olsak@cvut.cz}
}

\begin{document}

\maketitle

\begin{abstract}
We study how to make large language models~(LLMs) reliably generate working
Lean~4 tactic procedures from informal user specifications.
In a $2 \times 2$ experiment on the \texttt{limit\_auto} tactic---which
automatically proves \texttt{Tendsto~f~F~G} goals in Mathlib---we cross
milestone planning with live LSP compiler feedback.
Our sharpest finding is that \emph{LSP access to the live Lean kernel
matters more than planning alone}: one-shot generation with LSP feedback
(17 passing tests, 307 LOC) strictly outperforms a structured 5-milestone
plan without LSP (1 failing test, wrong Mathlib lemma names).
Combining both yields the strongest result (16 passing tests, 325 LOC,
zero errors after 9 milestones).
We characterise four error classes that only LSP feedback resolves---stale
API names, deprecated lemmas, elaboration-order pitfalls, and
\texttt{whnf}-unfolding surprises---and describe the specification-first
pipeline that underlies all successful conditions.
\end{abstract}

\keywords{Lean 4 \and tactic generation \and large language models \and
metaprogramming \and LLM-assisted formal methods \and LSP feedback}

%%% ============================================================
\section{Introduction}
%%% ============================================================

Interactive theorem provers such as Lean~4 have become important tools for
formalising mathematics and verifying software.  A key feature of these systems
is their extensibility through \emph{tactics}---procedures that automate proof
steps and can be reused across an entire development.  Writing a non-trivial
tactic requires deep knowledge of the \texttt{TacticM} monad, Lean's expression
representation, the Mathlib library API, and numerous elaboration subtleties
that are difficult to predict without a live compiler.

This paper investigates whether large language models can bridge this gap,
generating working Lean~4 tactics from informal user specifications.
Our main case study is a tactic we call \texttt{limit\_auto},
generated from the following user request (see \texttt{experiments/limit\_auto/spec.md}):
\begin{quote}
\textit{``A tactic that proves limit statements (tendsto goals) by analysing
the mathematical structure of the limit expression\ldots''}
\end{quote}
The resulting tactic automatically proves goals of the form \texttt{Tendsto~f~F~G}
by syntactically classifying the function~$f$ and selecting the appropriate Mathlib
lemma.  It requires non-trivial knowledge of filter notation, Mathlib
lemma names, and metaprogramming patterns---making it a good test case for our study.

Our original question was simply: \emph{can an LLM generate this tactic?}
Once we found that the answer depended critically on how the task was presented,
the question sharpened to \emph{which scaffolding strategies make generation successful}.
We compare four conditions in a $2 \times 2$ design:
\begin{itemize}
  \item the presence or absence of a formal \emph{milestone plan} generated
        before implementation; and
  \item the presence or absence of live \emph{LSP tool access} during
        implementation (real-time diagnostics, goal inspection, lemma search).
\end{itemize}

Our findings are that (1)~a raw one-shot API call without scaffolding produces
no usable output; (2)~a milestone plan alone helps structure the work but
cannot prevent wrong Mathlib API names from breaking the tactic; (3)~LSP
feedback alone (without a plan) produces a working one-shot implementation;
and (4)~combining both yields the most complete and readable result.
The sharpest result is that \textbf{LSP access to the live compiler is more
valuable than planning alone}: Condition~C (no plan, LSP) strictly dominates
Condition~B (plan, no LSP) on every metric.
By \emph{LSP tools} we mean the Lean Language Server Protocol tools available inside
Claude Code: real-time compiler diagnostics (\texttt{lean\_diagnostic\_messages}),
proof-state inspection (\texttt{lean\_goal}), and natural-language Mathlib
search (\texttt{lean\_leansearch}, \texttt{lean\_loogle}).
These give the model live feedback from the Lean kernel without requiring it to
run the compiler manually.

The paper also describes a pilot experiment on intuitionistic implication
tactics that motivated the specification-first pipeline used in the main study.

\paragraph{Contributions.}
\begin{itemize}
  \item A $2 \times 2$ empirical study of LLM tactic generation strategies.
  \item A characterisation of the error classes that LSP feedback uniquely resolves.
  \item A reusable pipeline: \textbf{spec} $\to$ \textbf{plan} $\to$
        \textbf{LSP-guided implementation}.
  \item The \texttt{limit\_auto} tactic as a concrete, verified artifact.
\end{itemize}

%%% ============================================================
\section{Background}
%%% ============================================================

\subsection{Lean 4 Metaprogramming}

Lean~4 provides a unified metaprogramming framework in which tactics are
written in Lean itself.  A tactic operates in the \texttt{TacticM}~monad, which
wraps \texttt{MetaM} (for unification and type inference) and
\texttt{ElabM} (for elaboration).  The key data types are:

\begin{itemize}
  \item \textbf{\texttt{Expr}}: the internal representation of Lean terms,
        types, and proofs.  Includes \texttt{.app}, \texttt{.lam},
        \texttt{.forallE}, \texttt{.lit}, etc.
  \item \textbf{\texttt{MVarId}}: a metavariable standing for an open proof
        obligation (a ``goal'').
  \item \textbf{\texttt{Syntax}}: the raw parse tree, used to quote and
        splice tactic expressions.
\end{itemize}

A tactic is registered via
\lstinline{syntax (name := myTac) "my_tac" : tactic} and
\lstinline{@[tactic myTac] def evalMyTac : Tactic := fun _ => do ...}.
The body inspects the current goal with \texttt{getMainGoal},
reads its type with \texttt{getType},
and can run sub-tactics via \texttt{evalTactic}.

\subsection{What Makes Tactic Generation Hard}

Generating a working tactic with an LLM is harder than generating ordinary
code for several reasons:

\begin{enumerate}
  \item \textbf{Mathlib API names.}  Mathlib contains tens of thousands of
        lemmas with non-obvious, versioned names.  Names change between
        releases (deprecation), and no LLM can reliably memorise all current
        identifiers.

  \item \textbf{Elaboration order.}  Lean's elaborator processes terms in a
        specific order.  Writing \texttt{exact f (by tac)} can cause
        \texttt{tac} to run before metavariables from the surrounding
        unification are resolved, producing errors that are invisible to
        static analysis.

  \item \textbf{Definition unfolding.}  \texttt{whnf} (weak-head normal form)
        aggressively unfolds definitions.  Calling \texttt{whnf} on a
        \texttt{Tendsto~f~F~G} expression silently unfolds \texttt{Filter.Tendsto}
        to \texttt{F.map~f~$\leq$~G}, making subsequent pattern matching fail.

  \item \textbf{Syntax node kinds.}  A multi-step tactic sequence has kind
        \texttt{tactic.seq}, not \texttt{tactic}.  Helpers typed as
        \texttt{TSyntax `tactic} reject sequences, causing type errors at
        the call site rather than the source of the mistake.
\end{enumerate}

\subsection{Prior Work}

LLM-based theorem proving has received significant attention.
LEGO-Prover~\cite{xia2024legoprover} accumulates reusable proof skills across sessions.
LeanDojo / ReProver~\cite{yang2024leandojo} retrieval-augments a language model
for Lean proof generation.
Hypertree Proof Search~\cite{lample2022hypertree} uses MCTS guided by a learned policy.
More recent systems---DeepSeek-Prover~\cite{deepseek2024prover},
AlphaProof~\cite{alphaproof2024}, and commercial efforts such as
Harmonic~\cite{harmonic2024}---have pushed the frontier of automated theorem proving
using reinforcement learning and large-scale Lean interaction.
All of these target \emph{proof generation}: finding sequences of \emph{existing} tactics.
\emph{Tactic synthesis}---generating new, reusable tactic procedures from
specifications---has received almost no attention in the literature,
despite being a practically important problem for library development.
Our work targets this gap.

Program synthesis from natural language specifications has been studied
extensively~\cite{gulwani2017program}, but the domain of
Lean metaprogramming combines a highly specialised API with strong
type-correctness requirements that differ qualitatively from general
code generation.

%%% ============================================================
\section{Pilot: Intuitionistic Implication Tactic}
%%% ============================================================

Before the main study, we ran a pilot experiment on a simpler target:
a tactic that proves goals of the form $P_1 \to \cdots \to P_n \to P_i$
(conclusion is one of the hypotheses).
We used a Julia program to enumerate provable propositional theorems
of this class (via Catalan and Bell numbers) as a test suite,
and explored three generation strategies with GPT-5.

\textbf{(1) Example-driven iteration}: feed generated theorems in batches
of increasing complexity, ask the model to update the tactic each time.
Produced 8 Lean files with no convergence---the tactic always matched
the current batch but did not generalise.

\textbf{(2) LLM-generated characterization}: ask the model to first
\emph{explain} the examples mathematically, then implement guided by
that characterization.  Converged in 3--5 files, but the inferred
characterization was narrower than the true class.

\textbf{(3) Researcher-written specification}: write the mathematical
specification ourselves---$\mathcal{F} = \{\varphi \mid \varphi \equiv
P_1 \to \cdots \to P_n \to Q,\; Q \in \{P_1,\ldots,P_n\}\}$, algorithm:
intros then assumption---then ask the model to implement.
Produced a correct, concise tactic in one session.

The key lesson: \emph{the model is a good code generator, but a poor
mathematician}.  Given a precise spec, it converges quickly; asked to
infer the class from examples, it does not.  This motivated the
specification-first pipeline used throughout the main study.
Full iteration logs are in \texttt{experiments/intuitionistic\_pilot/}.

%%% ============================================================
\section{The \texttt{limit\_auto} Study}
%%% ============================================================

\subsection{The Tactic}
\label{sec:tactic}

The tactic arising from the user request above is called \texttt{limit\_auto}
by the specification it generated for itself (Section~\ref{sec:spec}).
It automatically proves goals of the form \texttt{Tendsto~f~F~G} by
syntactically classifying the function~$f$ and dispatching to the appropriate Mathlib lemma.
It targets three pattern classes:

\begin{description}
  \item[Class~1 (Continuity)]
        $f$ is a composition of functions provably continuous by
        \texttt{continuity} or \texttt{fun\_prop}.
        Proof method: \texttt{Continuous.tendsto'} with the filter pair
        $(\mathcal{N}(a),\, \mathcal{N}(f(a)))$.

  \item[Class~2 (Exponential decay)]
        $f$ syntactically mentions \texttt{Real.exp}.
        Key lemmas: \texttt{Real.tendsto\_exp\_neg\_atTop\_nhds\_zero}
        ($\texttt{Tendsto}\;(x \mapsto e^{-x})\;\texttt{atTop}\;(\mathcal{N}\,0)$),
        and \texttt{Real.tendsto\_exp\_atBot}.

  \item[Class~3 (Reciprocal)]
        $f = x \mapsto x^{-1}$ in a normed division ring.
        Key lemmas: \texttt{Filter.tendsto\_inv$\text{\_}$0\_nhdsNE\_zero}
        and \texttt{Filter.tendsto\_inv$\text{\_}$0\_cobounded'}.
\end{description}

Additionally, the tactic performs a safety check that rejects syntactically
indeterminate forms ($t/t$, $(t{-}a)/(t{-}a)$, $t/0$, $0/0$)
before attempting any proof.

\subsection{The Specification}
\label{sec:spec}

The specification was generated by Claude~Sonnet from the one-line user request:
\begin{quote}
\textit{``A tactic that proves limit statements (tendsto goals) by analysing
the mathematical structure of the limit expression\ldots''}
\end{quote}
Before writing any code, the model was asked to produce:
(1)~a \emph{scope analysis} identifying the most feasible interpretation
given Lean's constraints;
(2)~a \emph{mathematical specification} naming the three classes with their
filter-pair conditions;
(3)~a \emph{test generation algorithm} describing how to systematically
enumerate provable \texttt{Tendsto} goals ordered by difficulty.

This specification served as the shared anchor for all four conditions:
every implementation was required to realise the same semantic contract.
The spec is stored at \texttt{experiments/limit\_auto/spec.md}.

\subsection{Study Design}

We study four strategies for \emph{generating} \texttt{limit\_auto} (Table~\ref{tab:design}).

\begin{table}[h]
\centering
\caption{$2 \times 2$ study design.}
\label{tab:design}
\begin{tabular}{lcc}
\toprule
                    & \textbf{No milestone plan} & \textbf{With milestone plan} \\
\midrule
\textbf{No LSP}     & A: \textit{one\_run/} (empty)    & B: \textit{plan\_without\_lsp/} \\
\textbf{With LSP}   & C: \textit{one\_shot\_with\_lsp/} & D: \textit{plan\_with\_lsp/} (best) \\
\bottomrule
\end{tabular}
\end{table}

An important methodological distinction separates the conditions.
\textbf{Condition A} was run via the automated pipeline (\texttt{pipeline/main.py}):
the model receives the spec via an API call, and a self-correction loop
automatically feeds compiler errors back for up to a fixed number of repair rounds.
\textbf{Conditions B, C, and D} were run interactively in Claude Code, where a human
researcher prompted the model to implement the tactic from the specification and,
where applicable, fed back the plan and previous milestone files at each step.
This distinction matters for interpreting Condition A's failure:
it is ambiguous whether the result reflects a limitation of the automated pipeline
scaffolding, the model, or both.

\textbf{Condition A} (no plan, no LSP) is the baseline: a single automated API call
with the spec as the only context, followed by a self-correction loop.
\textbf{Condition B} (plan, no LSP) is a Claude Code session in which the model
is given the spec and a milestone plan, and asked to implement each milestone
in sequence, with the previous milestone file and plan fed back as context at
each step, but no live compiler access.
\textbf{Condition C} (no plan, LSP) is a one-shot Claude Code session with full
LSP tool access throughout.
\textbf{Condition D} (plan + LSP) is a Claude Code session combining a 9-milestone
plan with LSP feedback at each step; this is the main positive condition.

\subsection{Results per Condition}
\label{sec:results}

Table~\ref{tab:results} summarises the four conditions quantitatively.

\begin{table}[h]
\centering
\caption{Quantitative comparison of the four conditions.
LOC = lines of code in the final tactic file.
Tests = theorem test cases in the final file.
Milestones = number of incremental implementation steps.
LSP calls = tool invocations recorded in the LSP log (Condition~D only; C used LSP interactively without a separate log).}
\label{tab:results}
\begin{tabular}{lcccccc}
\toprule
\textbf{Condition} & \textbf{Compiles} & \textbf{Tests pass} & \textbf{LOC} & \textbf{Tests} & \textbf{Milestones} & \textbf{LSP calls} \\
\midrule
A (no plan, no LSP) & No  & 0 / 0  & 0   & 0  & 1 & 0  \\
B (plan, no LSP)    & Yes & 8 / 9  & 160 & 9  & 5 & 0  \\
C (no plan, LSP)    & Yes & 17 / 17 & 307 & 17 & 1 & --- \\
D (plan, LSP)       & Yes & 16 / 16 & 325 & 16 & 9 & 34 \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Condition A.}
The automated pipeline produced no usable output: the model either declined to
generate metaprogramming code or produced a file that failed to compile, and the
self-correction loop did not converge within the repair budget.
As noted above, this failure is ambiguous: it may reflect the limitations of the
automated pipeline scaffolding (no interactive context, fixed repair budget),
the model's inability to synthesize metaprogramming code from a bare spec, or both.
To resolve this ambiguity, we ran a fifth condition informally called \textbf{Condition~E}:
the same task (no plan, no LSP) in an \emph{interactive} Claude Code session, eliminating
the pipeline as a confounding factor.
Condition~E produced 336 lines of code, but with a syntax error that prevented the tactic
from registering; all 13 test invocations failed with ``Tactic \texttt{limitAuto} has not
been implemented'' (0/13 pass).
This confirms that the failure in Condition~A is due to the \emph{absence of LSP feedback},
not the automated pipeline scaffolding.
Together, Conditions~A and~E establish the baseline: without live compiler access, neither
pipeline-driven nor interactive generation produces a working tactic.

\paragraph{Condition B.}
The milestone plan (five milestones) gave the model a structured decomposition
of the problem. The generated code compiled, but several tests failed at
runtime. The root cause was incorrect Mathlib lemma names: without a live
compiler, the model used names it ``knew'' from training data, including a
deprecated identifier
(\texttt{tendsto\_inv$\text{\_}$0\_nhdsWithin\_ne\_zero}, renamed to
\texttt{tendsto\_inv$\text{\_}$0\_nhdsNE\_zero}).
The plan helped with architecture but could not substitute for API verification.

\paragraph{Condition C.}
One-shot implementation with LSP access produced a working tactic with zero
compiler errors after iterative self-repair driven by \texttt{lean\_diagnostic\_messages}.
All tests pass.  The implementation is slightly less structured than
Condition~D (no explicit milestone decomposition) but is functionally correct.
This result shows that LSP access matters more than planning alone:
C (no plan, LSP) strictly dominates B (plan, no LSP).

\paragraph{Condition D.}
The nine-milestone plan with LSP feedback at each step produced the most
complete result: zero errors, the richest test suite, and the most readable
architecture with clearly separated helper functions, named pattern classes,
and a registered trace class (\texttt{limit\_auto}) for debugging.
The milestone plan provided \emph{error isolation}---bugs introduced in one
milestone were caught before they could propagate---and \emph{monotonic test growth}:
each milestone added new tests that were never broken by subsequent milestones.

\subsection{Concrete Failure Modes Fixed by LSP}
\label{sec:failures}

The LSP tools used in Condition~D include
\texttt{lean\_diagnostic\_messages} (compiler errors/warnings),
\texttt{lean\_goal} (proof state inspection),
\texttt{lean\_leansearch} (natural-language Mathlib search),
\texttt{lean\_loogle} (type-pattern search),
\texttt{lean\_multi\_attempt} (testing multiple tactics without editing),
and \texttt{lean\_local\_search} (local declaration search).
Together they caught four classes of errors that the LLM could not predict:

\paragraph{(1) \texttt{whnf} unfolds \texttt{Filter.Tendsto}.}
An early version called \texttt{whnf} on the goal before pattern-matching.
\texttt{whnf} aggressively unfolds \texttt{Filter.Tendsto} to its definition
$F.\texttt{map}\,f \leq G$, making \texttt{matchTendsto?} always return
\texttt{none}.  The tactic failed on every goal with the message
``goal is not of the form Tendsto~f~F~G''.
\texttt{lean\_goal} at the matching site revealed that the goal type had
already been unfolded.  Fix: remove the \texttt{whnf} call entirely and
match on \texttt{instantiateMVars ty} directly.

\paragraph{(2) \texttt{TSyntax `tactic} vs.\ \texttt{tactic.seq}.}
The helper \texttt{tryTac} was originally typed as
\texttt{tryTac : TSyntax `tactic $\to$ TacticM Bool}.
Multi-step sequences such as
\texttt{apply Continuous.tendsto' <;> first | continuity | norm\_num}
have syntax node kind \texttt{tactic.seq}, not \texttt{tactic},
causing a type error at every call site that used a compound tactic.
\texttt{lean\_diagnostic\_messages} identified the kind mismatch.
Fix: change the argument type to \texttt{Syntax}
(which \texttt{evalTactic} accepts directly).

\paragraph{(3) \texttt{simp} inside \texttt{exact~f~(by~simp)}.}
For continuity goals with numeric targets (e.g., $2+1=3$),
the natural proof is \texttt{exact Continuous.tendsto' (by continuity) \_ \_ (by simp)}.
However, when \texttt{simp} runs inside a term elaboration context and makes
no progress, it writes an error to the message log \emph{before} throwing an
exception---so the error persists even when \texttt{tryTac} catches it.
This produced spurious diagnostics even on successful proofs.
Fix: replace the term-mode form with
\texttt{apply Continuous.tendsto' <;> first | continuity | norm\_num | ring | rfl | simp},
which runs tactics in sequence on subgoals and avoids the elaboration issue.

\paragraph{(4) Deprecated lemma names.}
\texttt{lean\_leansearch} in Milestone~6 found the cobounded inverse lemma as
\texttt{Filter.tendsto\_inv$\text{\_}$0\_nhdsWithin\_ne\_zero}.  The compiler
immediately flagged a deprecation warning:
the current name is \texttt{Filter.tendsto\_inv$\text{\_}$0\_nhdsNE\_zero}.
Without LSP access (Condition~B), the model used the old name, causing
the reciprocal tests to fail at runtime.

\bigskip

A summary of the complete LSP tool usage in Condition~D is given in
Table~\ref{tab:lsp}.

\begin{table}[h]
\centering
\caption{LSP tools used in Condition~D (9-milestone implementation).}
\label{tab:lsp}
\begin{tabular}{llr}
\toprule
\textbf{Tool} & \textbf{Purpose} & \textbf{Uses} \\
\midrule
\texttt{lean\_diagnostic\_messages} & Compiler errors/warnings at each milestone & 18 \\
\texttt{lean\_leansearch}           & Natural-language Mathlib lemma search     & 6  \\
\texttt{lean\_loogle}               & Type-pattern Mathlib search               & 3  \\
\texttt{lean\_multi\_attempt}       & Test multiple tactics without editing     & 2  \\
\texttt{lean\_goal}                 & Proof state inspection                    & 2  \\
\texttt{lean\_local\_search}        & Local declaration search                  & 4  \\
\texttt{lean\_hover\_info}          & Type signature of an identifier           & 1  \\
\bottomrule
\end{tabular}
\end{table}

\noindent
Table~\ref{tab:lsp} is derived from \texttt{experiments/limit\_auto/plan\_with\_lsp/lsp\_log.md},
which records every LSP tool call made during Condition~D.
The model used was Claude~Sonnet~4.6, running inside Claude Code
with a \texttt{CLAUDE.md} project file specifying the Lean toolchain and Mathlib imports.
All four condition outputs are archived in the repository at
\url{https://github.com/Sfgangloff/tactics_generation}
under \texttt{experiments/limit\_auto/}.

%%% ============================================================
\section{Discussion}
%%% ============================================================

\paragraph{The specification is the hardest step.}
All four conditions shared the same specification, yet only the implementations
with LSP access produced correct code.  This suggests that a bad specification
is fatal regardless of scaffolding, while a good specification is necessary but
not sufficient.  Investing effort in specification quality---scope analysis,
precise class definitions, provability conditions---is the highest-leverage
action in the pipeline.

\paragraph{LSP feedback is essential for API correctness.}
No LLM can reliably memorise the current state of Mathlib's tens of thousands
of lemma names, especially as names are deprecated and renamed across releases.
The results confirm this: Condition~B failed not from logical mistakes but from
stale API knowledge.  A live compiler acts as a ground truth that the model
cannot replicate internally.

\paragraph{Plans provide error isolation, not API knowledge.}
The milestone plan in Condition~D did not contribute to correct lemma names---
that was the role of \texttt{lean\_leansearch} and \texttt{lean\_loogle}.
What the plan contributed was \emph{modularity}: each milestone had a narrow
scope, so errors introduced at one step were caught before they could obscure
errors at later steps.  The plan also ensured that each milestone added new
tests, giving a concrete definition of ``done'' at each step.

\paragraph{Condition~C beats Condition~B.}
This ordering---one-shot with LSP access beats planned without LSP---is perhaps
the sharpest result of the study.  It implies that tool access to the live
compiler is more valuable than decomposing the problem into milestones, at
least for a tactic of this size.  For larger tactics, the balance may shift.

%%% ============================================================
\section{Related Work}
%%% ============================================================

\paragraph{LLM-based proof generation.}
LEGO-Prover~\cite{xia2024legoprover} uses an LLM to generate and accumulate
reusable proof ``skills'' across a session.  Copilot for Lean~\cite{han2022proof}
and ReProver~\cite{yang2024leandojo} train models to predict the next tactic
in a proof.  Hypertree Proof Search~\cite{lample2022hypertree} frames proof
generation as MCTS with a learned policy.  These works target \emph{proof
generation}---finding sequences of \emph{existing} tactics---whereas our work
targets \emph{tactic synthesis}: generating new, reusable tactic procedures.

\paragraph{Program synthesis from natural language.}
The literature on program synthesis from specifications is extensive;
see~\cite{gulwani2017program} for a survey.  LLM-based code generation
(Codex~\cite{chen2021codex}, AlphaCode~\cite{li2022competition}) has
demonstrated strong performance on standard benchmarks.
Lean metaprogramming is qualitatively harder because correctness requires
type-level compliance with a complex API.

\paragraph{Tool-augmented LLMs.}
Using tools (compilers, search engines, REPL environments) to augment LLM
reasoning has become standard practice~\cite{schick2023toolformer}.
Our work is an instance of this paradigm applied to Lean metaprogramming,
with the Lean LSP server as the primary tool.
Closest in spirit is the use of REPL feedback in
AlphaCode~2~\cite{alphacode2} and in interactive proof
assistants~\cite{first2023baldur}.
SatLM~\cite{ye2023satlm} makes the connection between declarative specification
and external verification explicit: the LLM generates a task specification and
delegates solving to an automated prover.  Our approach mirrors this split---
the \texttt{analyze\_request} stage produces a specification and the Lean kernel
acts as the verifier---applied to the harder problem of metaprogramming synthesis.

\paragraph{Prompting strategies for structured generation.}
Several prompting frameworks parallel our scaffolding choices.
\emph{Least-to-Most Prompting}~\cite{zhou2022least} and
\emph{Plan-and-Solve Prompting}~\cite{wang2023planandsolve} decompose a complex
problem into ordered subproblems solved sequentially---exactly the role our milestone
plan plays in Conditions~B and~D.
\emph{Problem Elaboration Prompting}~\cite{liao2024leap} (``look before you leap'')
proposes clarifying problem context before reasoning, mirroring our
specification-first approach: the \texttt{analyze\_request} stage elaborates the
user request into a precise mathematical spec before any code is generated.
\emph{Progressive-Hint Prompting}~\cite{zheng2023php} uses previously generated
outputs as hints to guide subsequent attempts; the self-correction loop in
Conditions~A and~B instantiates this pattern, feeding compiler errors back as
hints.
Finally, \emph{Cumulative Reasoning}~\cite{zhang2023cumulative} formalises a
Proposer--Verifier--Reporter framework in which a Proposer generates steps, a
Verifier checks them, and a Reporter decides when the solution is complete.
Conditions~C and~D realise this architecture: Claude Code acts as the Proposer,
the Lean LSP tools as the Verifier, and the model's own judgement as the Reporter.

%%% ============================================================
\section{Conclusion}
%%% ============================================================

We have presented an empirical $2 \times 2$ study of LLM strategies for
Lean~4 tactic generation, using the \texttt{limit\_auto} tactic as a
test case.  The main findings are:
\begin{enumerate}
  \item Raw one-shot generation is not viable.
  \item A formal specification anchors all conditions and is the prerequisite
        for any successful generation.
  \item LSP feedback fixes errors (wrong API names, elaboration pitfalls)
        that LLMs cannot predict statically.
  \item Milestone planning and LSP feedback are complementary, and their
        combination produces the most complete result.
\end{enumerate}

\paragraph{Future work.}
As a preliminary step beyond \texttt{limit\_auto}, we ran the plan + LSP strategy
(Condition D) on a second tactic: \texttt{decide\_list\_theory}, which automatically
decides list-theoretic propositions (membership, equality, sublist).
The generation followed the same interactive Claude Code methodology as Condition D
(specification generated by the pipeline, milestone plan written by the researcher,
milestones implemented one by one with LSP access).
This provides a second data point suggesting the strategy is not specific to
\texttt{limit\_auto}, though a systematic multi-tactic study remains future work.
The repository also includes 12 draft specifications generated by the pipeline's
\texttt{analyze\_request} stage as candidates for future implementation experiments.

The most important next step is to run the same $2 \times 2$ study on additional
tactics to test whether the ordering of conditions is robust.  Longer-term, we hope
to build a benchmark suite of ``tactics from specifications'' that the community can
use to evaluate LLM-assisted metaprogramming systems, and to study whether the
approach generalises to tactics requiring deeper mathematical reasoning than
pattern matching.

%%% ============================================================
\bibliographystyle{splncs04}
\begin{thebibliography}{9}

\bibitem{xia2024legoprover}
Huajian Xia et al.
\textit{LEGO-Prover: Neural Theorem Proving with Growing Libraries}.
ICLR 2024.

\bibitem{han2022proof}
Jesse Michael Han, Jason Rute, Yuhuai Wu, Edward Ayers, Stanislas Polu.
\textit{Proof Artifact Co-Training}.
ICLR 2022.

\bibitem{yang2024leandojo}
Kaiyu Yang et al.
\textit{LeanDojo: Theorem Proving with Retrieval-Augmented Language Models}.
NeurIPS 2023.

\bibitem{lample2022hypertree}
Guillaume Lample et al.
\textit{HyperTree Proof Search for Neural Theorem Proving}.
NeurIPS 2022.

\bibitem{gulwani2017program}
Sumit Gulwani, Oleksandr Polozov, Rishabh Singh.
\textit{Program Synthesis}.
Foundations and Trends in Programming Languages, 2017.

\bibitem{chen2021codex}
Mark Chen et al.
\textit{Evaluating Large Language Models Trained on Code}.
arXiv:2107.03374, 2021.

\bibitem{li2022competition}
Yujia Li et al.
\textit{Competition-Level Code Generation with AlphaCode}.
Science 378(6624), 2022.

\bibitem{schick2023toolformer}
Timo Schick et al.
\textit{Toolformer: Language Models Can Teach Themselves to Use Tools}.
NeurIPS 2023.

\bibitem{alphacode2}
Google DeepMind.
\textit{AlphaCode 2 Technical Report}.
2023.

\bibitem{first2023baldur}
Emily First, Markus Rabe, Talia Ringer, Yuriy Brun.
\textit{Baldur: Whole-Proof Generation and Repair with Large Language Models}.
ESEC/FSE 2023.

\bibitem{deepseek2024prover}
DeepSeek-AI.
\textit{DeepSeek-Prover-V1.5: Harnessing Proof Assistant Feedback for
Reinforcement Learning and Monte-Carlo Tree Search}.
arXiv:2408.08152, 2024.

\bibitem{alphaproof2024}
Google DeepMind.
\textit{AI achieves silver-medal standard solving International Mathematical
Olympiad problems} (AlphaProof).
Technical Report, 2024.

\bibitem{harmonic2024}
Harmonic.
\textit{Harmonic: Formal Mathematics with AI}.
\url{https://harmonic.fun}, 2024.

\bibitem{zhou2022least}
Denny Zhou et al.
\textit{Least-to-Most Prompting Enables Complex Reasoning in Large Language Models}.
ICLR 2023. arXiv:2205.10625.

\bibitem{wang2023planandsolve}
Lei Wang et al.
\textit{Plan-and-Solve Prompting: Improving Zero-Shot Chain-of-Thought Reasoning by Large Language Models}.
ACL 2023. arXiv:2305.04091.

\bibitem{liao2024leap}
Haoran Liao et al.
\textit{Look Before You Leap: Problem Elaboration Prompting Improves Mathematical Reasoning in Large Language Models}.
arXiv:2402.15764, 2024.

\bibitem{zheng2023php}
Chuanyang Zheng et al.
\textit{Progressive-Hint Prompting Improves Reasoning in Large Language Models}.
arXiv:2304.09797, 2023.

\bibitem{zhang2023cumulative}
Yifan Zhang et al.
\textit{Cumulative Reasoning with Large Language Models}.
arXiv:2308.04371, 2023.

\bibitem{ye2023satlm}
Xi Ye et al.
\textit{SatLM: Satisfiability-Aided Language Models Using Declarative Prompting}.
NeurIPS 2023. arXiv:2305.09656.

\bibitem{lean4}
Leonardo de Moura and Sebastian Ullrich.
\textit{The Lean~4 Theorem Prover and Programming Language}.
CADE 2021.

\bibitem{mathlib}
The Mathlib Community.
\textit{The Lean Mathematical Library}.
CPP 2020.

\end{thebibliography}

\end{document}
